---
output: github_document
bibliography: vignettes/references.bib
link-citations: true
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```
[![Travis-CI Build Status](https://travis-ci.org/sbfnk/bpmodels.svg?branch=master)](https://travis-ci.org/sbfnk/bpmodels)
[![Appveyor Build Status](https://ci.appveyor.com/api/projects/status/y37i8x0wo9o8s2wf?svg=true)](https://ci.appveyor.com/project/sbfnk/bpmodels)
[![codecov](https://codecov.io/github/sbfnk/bpmodels/branch/master/graphs/badge.svg)](https://codecov.io/github/sbfnk/bpmodels) 
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

`bpmodels` is an R package to simulate and analyse the size and length of branching processes with a given offspring distribution.

# Installation
The latest development version of the `bpmodels` package can be installed via

```{r eval=FALSE}
devtools::install_github('sbfnk/bpmodels')
```

Please note that the 'bpmodels' project is released with a [Contributor Code of Conduct](CODE_OF_CONDUCT.md). By contributing to this project, you agree to abide by its terms.
