---
output: github_document
bibliography: vignettes/references.json
link-citations: true
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = file.path("man", "figures", "README-"),
  out.width = "100%"
)
```

# _bpmodels_: Methods for analysing the size and length of transmission chains from branching process models

<!-- badges: start -->
![GitHub R package version](https://img.shields.io/github/r-package/v/epiverse-trace/bpmodels)
[![R-CMD-check](https://github.com/epiverse-trace/bpmodels/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/epiverse-trace/bpmodels/actions/workflows/R-CMD-check.yaml)
[![codecov](https://codecov.io/github/epiverse-trace/bpmodels/branch/main/graphs/badge.svg)](https://codecov.io/github/epiverse-trace/bpmodels) 
![GitHub contributors](https://img.shields.io/github/contributors/epiverse-trace/bpmodels)
[![License: MIT](https://img.shields.io/badge/License-MIT-yellow.svg)](https://opensource.org/licenses/MIT)
<!-- badges: end -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

_bpmodels_ is an R package to simulate and analyse the size and length of 
branching processes with a given offspring distribution. These models are often 
used in infectious disease epidemiology, where the chains represent chains of
transmission, and the offspring distribution represents the distribution of 
secondary infections caused by an infected individual.

# Installation

The latest development version of the _bpmodels_ package can be installed via

```{r include=TRUE,eval=FALSE}
# check whether {pak} is installed
if(!require("pak")) install.packages("pak")
pak::pkg_install("epiverse-trace/bpmodels")
```

To load the package, use

```{r eval=TRUE}
library("bpmodels")
```

# Core functionality

_bpmodels_ provides three functions: 

* `chain_ll()`: calculates the likelihoods of observing a vector of chains 
  of given sizes or lengths.

* `chain_sim()`: simulates transmission chains until all chains die out.

* `chain_sim_susc()`: simulates transmission chains until a specified initial 
  susceptible pool runs out.


See the ["introduction vignette"](bpmodels) for a detailed illustration of 
each function.

## Package vignettes

Specific use cases of _bpmodels_ can be found in 
the [online documentation as package vignettes](https://epiverse-trace.github.io/bpmodels/), under "Articles".

## Reporting bugs 

To report a bug please open an [issue](https://github.com/epiverse-trace/bpmodels/issues/new/choose).

## Contribute

We welcome contributions to enhance the package's functionalities. If you 
wish to do so, please follow the [package contributing guide](https://github.com/epiverse-trace/bpmodels/blob/main/.github/CONTRIBUTING.md).

## Code of conduct

Please note that the _bpmodels_ project is released with a [Contributor Code of Conduct](https://github.com/epiverse-trace/.github/blob/main/CODE_OF_CONDUCT.md). 
By contributing to this project, you agree to abide by its terms.

## Citing this package

```{r message=FALSE, warning=FALSE}
citation("bpmodels")
```
