---
output: github_document
bibliography: vignettes/references.json
link-citations: true
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = file.path("man", "figures", "README-"),
  out.width = "100%"
)
```

# _bpmodels_: Methods for analysing the size and length of transmission chains from branching process models

<!-- badges: start -->
![GitHub R package version](https://img.shields.io/github/r-package/v/epiverse-trace/bpmodels)
[![R-CMD-check](https://github.com/epiverse-trace/bpmodels/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/epiverse-trace/bpmodels/actions/workflows/R-CMD-check.yaml)
[![codecov](https://codecov.io/github/epiverse-trace/bpmodels/branch/main/graphs/badge.svg)](https://codecov.io/github/epiverse-trace/bpmodels) 
![GitHub contributors](https://img.shields.io/github/contributors/epiverse-trace/bpmodels)
[![License: MIT](https://img.shields.io/badge/License-MIT-yellow.svg)](https://opensource.org/licenses/MIT)
<!-- badges: end -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

_bpmodels_ is an R package to simulate and analyse the size and length of 
branching processes with a given offspring distribution. These models are often 
used in infectious disease epidemiology, where the chains represent chains of
transmission, and the offspring distribution represents the distribution of 
secondary infections caused by an infected individual.

# Installation

The latest development version of the _bpmodels_ package can be installed via

```{r eval=FALSE}
# check whether {pak} is installed
if (!require("pak")) install.packages("pak")
pak::pkg_install("epiverse-trace/bpmodels")
```

To load the package, use

```{r eval=TRUE}
library("bpmodels")
```

# Core functionality

_bpmodels_ provides three main functions: 

`chain_ll()`: calculates the likelihoods of observing a vector of chains 
  of given sizes or lengths.

Here is a quick example of estimating the loglikelihood of an observed chain:
```{r}
# example of observed chain sizes
chain_sizes <- c(1, 2, 3, 4) 
# estimate loglikelihood of the observed chain sizes
chain_ll_eg <- chain_ll(chain_sizes, "pois", "size", lambda = 0.5)
chain_ll_eg
```


`chain_sim()`: simulates transmission chains until all chains stop producing
  offspring.

Below is a quick example where we simulate the chain sizes of $5$ chains with 
a poisson offspring with mean, $\text{lambda} = 0.5$:
```{r}
set.seed(123)

chain_sim_eg <- chain_sim(n = 5, offspring = "pois", stat = "size", 
                          lambda = 0.5, tree = TRUE)

head(chain_sim_eg)
```

`chain_sim_susc()`: simulates transmission chains from a specified population
  size with pre-existing immunity until the susceptible pool runs out.
  
Below is a quick example where we simulate chains with a poisson 
offspring with mean, $\text{lambda} = 0.5$, and serial interval of $3$:
```{r}
set.seed(1234)

chain_sim_susc_eg <- chain_sim_susc(pop = 1000, "pois",
                                    mn_offspring = 0.5,
                                    serial = function(x) 3
                                    )

head(chain_sim_susc_eg)
```

See the ["Get started vignette"](https://epiverse-trace.github.io/bpmodels/articles/bpmodels.html) for a detailed illustration of 
each function.

## Package vignettes

Specific use cases of _bpmodels_ can be found in 
the [online documentation as package vignettes](https://epiverse-trace.github.io/bpmodels/), under "Articles".

## Reporting bugs 

To report a bug please open an [issue](https://github.com/epiverse-trace/bpmodels/issues/new/choose).

## Contribute

We welcome contributions to enhance the package's functionalities. If you 
wish to do so, please follow the [package contributing guide](https://github.com/epiverse-trace/bpmodels/blob/main/.github/CONTRIBUTING.md).

## Code of conduct

Please note that the _bpmodels_ project is released with a [Contributor Code of Conduct](https://github.com/epiverse-trace/.github/blob/main/CODE_OF_CONDUCT.md). 
By contributing to this project, you agree to abide by its terms.

## Citing this package

```{r message=FALSE, warning=FALSE}
citation("bpmodels")
```
