---
title: "Analysing infectious disease transmission chains with bpmodels"
author: "Sebastian Funk, James Azam"
output:
  bookdown::html_vignette2:
    fig_caption: yes
    code_folding: show
pkgdown:
  as_is: true
bibliography: references.json
link-citations: true
vignette: >
  %\VignetteEncoding{UTF-8}
  %\VignetteIndexEntry{Analysing infectious disease transmission chains with bpmodels}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

# Note

> `{bpmodels}` is now *retired and will no longer be maintained*. We recommend using [`{epichains}`](https://github.com/epiverse-trace/epichains) instead. If you need help converting your code to use `{epichains}`, please [open a discussion on epichains](https://github.com/epiverse-trace/epichains/discussions).

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message = FALSE,
                      warning = FALSE,
                      collapse = TRUE,
                      comment = "#>"
                      )

```

_bpmodels_ provides methods to analyse and simulate the size and length
of branching processes with an arbitrary offspring distribution. These
can be used, for example, to analyse the distribution of chain sizes
or length of infectious disease outbreaks, as discussed in @farrington2003 and 
@blumberg2013.
    
# Quick start

```{r}
library("bpmodels")
```

## `chain_ll()`: calculate log-likelihoods

The `chain_ll()` function calculates the log-likelihood of a distribution of 
chain sizes or lengths given an offspring distribution and its associated 
parameters. 

For example, if we have observed a distribution of chains of sizes 
$1, 1, 4, 7$, we can 
calculate the log-likelihood of this observed chain by assuming the offspring 
per generation is Poisson distributed with a mean number (which can 
be interpreted as the reproduction number, $R_0$) of $0.5$. 

To do this, we run 

```{r}
chain_sizes <- c(1, 1, 4, 7) # example of observed chain sizes
chain_ll(x = chain_sizes, offspring = "pois", stat = "size", lambda = 0.5)
```

The first argument of `chain_ll()` is the chain size (or length, in number of 
generations that a chain lasted) distribution to 
analyse. 

The second argument, `offspring`, specifies the offspring 
distribution. This is specified as a string of the name of the function used to generate random offspring. 
It can be any probability distribution implemented in `R`, that is, one that 
has a corresponding function for generating random numbers beginning with the 
letter `r`. In the case of the example above, since random Poisson numbers are 
generated in `R` using a function called `rpois()`, the string to pass to the 
`offspring` argument is `"pois"`.

The third argument, `stat`, determines whether to analyse chain sizes 
(`"size"`, the default if this argument is not specified) or lengths 
(`"length"`). Lastly, any named arguments not recognised by `chain_ll()` 
are interpreted as parameters of the corresponding probability distribution, 
here `lambda = 0.5` as the mean of the Poisson distribution (see the `R` help 
page for the [Poisson distribution](https://stat.ethz.ch/R-manual/R-devel/library/stats/html/Poisson.html) for more information). 

### Imperfect observations

By default, `chain_ll()` assumes perfect observation, where `obs_prob = 1` 
(See `?chain_ll`), meaning that all transmission events are observed and 
recorded in the data. If observations are imperfect, `chain_ll()` provides 
the argument, `obs_prob`, for specifying the probability of observation. 
This probability is used to determine the likelihood of observing the specified
chain sizes or lengths. In the case of imperfect observation, true chain sizes 
or lengths are simulated repeatedly (the number of times given by the 
`nsim_obs` argument), and the likelihood calculated for each of 
these simulations. 

For example, if the probability of observing each case is `obs_prob = 0.30`, 
we use

```{r}
chain_sizes <- c(1, 1, 4, 7) # example of observed chain sizes
ll <- chain_ll(chain_sizes, "pois", "size", obs_prob = 0.3, lambda = 0.5,
               nsim_obs = 10)
ll
```

This returns `10` likelihood values (because `nsim_obs = 10`), which can be 
averaged to come up with an overall likelihood estimate.

To find out about the usage of the `chain_ll()` function, you can run `?chain_ll` 
to access its `R` help file.

### How `chain_ll()` works

If the probability distribution of chain sizes or lengths has an analytical 
solution, this will be used. `chain_ll()` currently supports the Poisson and 
negative binomial size distribution and the Poisson and geometric length 
distribution. 

If an analytical solution does not exist, simulations are used to approximate 
this probability distributions ([using a linear approximation to the cumulative 
distribution](https://en.wikipedia.org/wiki/Empirical_distribution_function) 
for unobserved sizes/lengths). In that case, an extra argument `nsim_offspring` 
must be passed to `chain_ll()` to specify the number of simulations to be 
used for this approximation. 

For example, to get offspring drawn from a binomial distribution with 
probability `prob = 0.5`, we run

```{r}
set.seed(12)
chain_sizes <- c(1, 1, 4, 7) # example of observed chain sizes
chain_ll(chain_sizes, "binom", "size", size = 1, prob = 0.5,
         nsim_offspring = 100)
```

## `chain_sim()`: simulate transmission chains with branching processes

To simulate a branching process, we use the `chain_sim()` function. This 
function follows the same syntax as `chain_ll()`. Note that `chain_sim()` is
stochastic, so a seed needs to be set to ensure that the results can be
reproduced.

Below, we are simulating $5$ chains, assuming the offspring are generated using
a Poisson distribution with mean, `lambda = 0.5`. By default, `chain_sim()` 
returns a vector of chain sizes/lengths. If we instead want to return 
a tree of infectees and infectors, we need to specify a function for 
the serial interval and set `tree = TRUE` (see next section).

```{r}
set.seed(12)
chain_sim(n = 5, offspring = "pois", stat = "size", lambda = 0.5)
```

### Simulating trees

To simulate a tree of transmission chains, we specify the serial interval 
generation function (`serial_interval()`) and set `tree = TRUE` as follows:

```{r}
set.seed(12)
serial_interval <- function(n) {
  rlnorm(n, meanlog = 0.58, sdlog = 1.58)
}
chains_df <- chain_sim(
  n = 5, offspring = "pois", lambda = 0.5, stat = "length",
  infinite = 100, serial = serial_interval, tree = TRUE
)
head(chains_df)
```

## References
