---
title: "Analysing chain statistics using branching process models"
author: "Sebastian Funk"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Analysing chain statistics using branching process models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
library('knitr')
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

[bpmodels](https://github.com/sbfnk/bpmodels) is an `R` package to analyse and simulate the size and length of branching processes with an arbitrary offspring distribution. These can be used, for example, to analyse the distribution of chain sizes or length of infectious disease outbreaks.

# Usage

To load the package, use
```{r eval=FALSE}
library('bpmodels')
```
```{r echo=FALSE}
suppressWarnings(library('bpmodels'))
```

At the heart of the `bpmodels` package are the `chains_ll` and `chains_sim` functions. The `chains_ll` function calculates the log-likelihood of a distribution of chain sizes or lengths given an offspring distribution and associated parameters. For example, to get the log-likelihood for a given observed distribution of chain sizes assuming a mean number of 0.5 Poisson-distributed offspring per generation, use

```{r}
chain_sizes <- c(1,1,4,7) # example of observed chain sizes
chain_ll(chain_sizes, rpois, "size", lambda=0.5)
```

The first argument of `chain_ll` is the size (or length) distribution to analyse. The second argument (called `offspring`) specifies the offspring distribution. This is given as a the function used to generate random offspring. It can be any probability distribution implemented in R, that is, one that has a corresponding function for generating random numbers beginning with the letter `r`. In the case of the example above, since random Poisson numbers are generated in R using a function called `rpois`, this is the function to pass as the `offspring` argument.

The third argument (called `stat`) determines whether to analyse chain sizes ("size", the default if this argument is not specified) or lengths ("length"). Lastly, any named arguments not recognised by `chain_ll` are interpreted as parameters of the corresponding probability distribution, here `lambda=0.5` as the mean of the Poisson distribution (see the R help page for the Poisson distribution for more information).

You can use the `R` help to find out about usage of the `chains_ll` function,

```{r eval=FALSE}
?chains_ll
```

To simulate from a branching process, use the `chain_sim` function, which follows the same syntax as the `chain_ll` function:

```{r}
chain_sim(n=5, rpois, "size", lambda=0.5)
```

# Methodology

If the probability distribution of chain sizes or lengths has an analytical solution, this will be used (size distribution: Poisson and negative binomial; length distribution: Poisson and geometric). If not, simulations are used to approximate this probability distributions (using a linear approximation to the cumulative distribution for unobserved sizes/lengths), requiring an additional parameter `nsim_offspring` for the number of simulations to be used for this approximation. For example, to get offspring drawn from a binomial distribution with probability `p=0.5`.

```{r}
chain_ll(chain_sizes, rbinom, "size", size=1, prob=0.5, nsim_offspring=100)
```

# Imperfect observations

The `chain_ll` function has an `obs_prob` parameter that can be used to determine the likelihood if observations are imperfect. This only works when analysing chain sizes (`stat="size"`). In that case, true chain sizes are simulated repeatedly (the number of times given by the `nsim_obs` argument) and the likelihood calculated for each of these simulations. For example, if the probability of observing each case is 30%, use

```{r}
ll <- chain_ll(chain_sizes, rpois, "size", obs_prob = 0.3, lambda=0.5, nsim_obs=10)
summary(ll)
```

This returns `nsim_obs=10` likelihood values which can be averaged to come up with an overall likelihood estimate.

# References

* Farrington, C.P., Kanaan, M.N. and Gay, N.J. (2003). [Branching process models for surveillance of infectious diseases controlled by mass vaccination](https://doi.org/10.1093/biostatistics/4.2.279).
* Blumberg, S. and Lloyd-Smith, J.O. (2013). [Comparing methods for estimating R0 from the size distribution of subcritical transmission chains](https://doi.org/10.1016/j.epidem.2013.05.002).
