---
title: "Analysing chain statistics using branching process models"
author: "Sebastian Funk"
output:
  bookdown::html_vignette2:
    fig_caption: yes
    code_folding: show
pkgdown:
  as_is: true
bibliography: references.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{Analysing chain statistics using branching process models}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---


```{r setup, include = FALSE}
library('knitr')
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

[bpmodels](https://github.com/sbfnk/bpmodels) is an `R` package to simulate and analyse the size and length of branching processes with a given offspring distribution. These can be used, for example, to analyse the distribution of chain sizes or length of infectious disease outbreaks.

# Quick start

To load the package, use
```{r eval=FALSE}
library('bpmodels')
```
```{r echo=FALSE}
suppressWarnings(library('bpmodels'))
set.seed(13)
```

At the heart of the package are the `chains_ll()` and `chains_sim()` functions. 

## Calculating log-likelihoods

The `chains_ll()` function calculates the log-likelihood of a distribution of chain sizes or lengths given an offspring distribution and its associated parameters. 

If we have observed a distribution of chains of sizes $1, 1, 4, 7$, we can calculate the log-likelihood of this observed chain by assuming the offspring per generation is Poisson distributed with a mean number of 0.5. 

To do this, we run 

```{r}
chain_sizes <- c(1, 1, 4, 7) # example of observed chain sizes
chain_ll(chain_sizes, "pois", "size", lambda = 0.5)
```

The first argument of `chain_ll()` is the size (or length) distribution to analyse. The second argument (called `offspring`) specifies the offspring distribution. This is given as a function used to generate random offspring. It can be any probability distribution implemented in `R`, that is, one that has a corresponding function for generating random numbers beginning with the letter `r`. In the case of the example above, since random Poisson numbers are generated in `R` using a function called `rpois()`, the string to pass to the `offspring` argument is `"pois"`.

The third argument (called `stat`) determines whether to analyse chain sizes (`"size"`, the default if this argument is not specified) or lengths (`"length"`). Lastly, any named arguments not recognised by `chain_ll()` are interpreted as parameters of the corresponding probability distribution, here `lambda = 0.5` as the mean of the Poisson distribution (see the `R` help page for the [Poisson distribution](https://stat.ethz.ch/R-manual/R-devel/library/stats/html/Poisson.html) for more information).

To find out about usage of the `chains_ll()` function, you can use the `R` help file

```{r eval=FALSE}
?chains_ll
```

## Simulating branching processes

To simulate a branching process, we use the `chain_sim()` function. This function follows the same syntax as `chain_ll()`, that is:

```{r}
chain_sim(n = 5, "pois", "size", lambda = 0.5)
```

# Methodology

If the probability distribution of chain sizes or lengths has an analytical solution, this will be used (size distribution: Poisson and negative binomial; length distribution: Poisson and geometric). 

If an analytical solution does not exist, simulations are used to approximate this probability distributions (using a linear approximation to the cumulative distribution for unobserved sizes/lengths). The argument `nsim_offspring` is used to specify the number of simulations to be used for this approximation. 

For example, to get offspring drawn from a binomial distribution with probability `prob = 0.5`, we run

```{r}
chain_ll(chain_sizes, "binom", "size", size = 1, prob = 0.5, nsim_offspring = 100)
```

# Imperfect observations

If observations are imperfect, the `chain_ll()` function has an `obs_prob` argument that can be used to determine the likelihood. In that case, true chain sizes or lengths are simulated repeatedly (the number of times given by the `nsim_obs` argument), and the likelihood calculated for each of these simulations. 

For example, if the probability of observing each case is $30%$, we use

```{r}
ll <- chain_ll(chain_sizes, "pois", "size", obs_prob = 0.3, lambda = 0.5, nsim_obs = 10)
summary(ll)
```

This returns `nsim_obs = 10` likelihood values which can be averaged to come up with an overall likelihood estimate.

# References

* Farrington, C.P., Kanaan, M.N. and Gay, N.J. (2003). [Branching process models for surveillance of infectious diseases controlled by mass vaccination](https://doi.org/10.1093/biostatistics/4.2.279).
* Blumberg, S. and Lloyd-Smith, J.O. (2013). [Comparing methods for estimating R0 from the size distribution of subcritical transmission chains](https://doi.org/10.1016/j.epidem.2013.05.002).
