---
title: "Projecting future disease incidence given early outbreak data"
author: "James Azam, Sebastian Funk"
date: '2023-01-13'
output: html_document
bibliography: vignettes/references.bib 
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

library("bpmodels")
library("readr")
library("lubridate")
library('dplyr')
library('ggplot2')
```

## Description
Branching processes can be used to project disease incidence provided we have some 
information on the distribution of times between successive cases (serial interval), 
and the distribution of secondary cases produced by a single individual (offspring 
distribution). Such simulations can be achieved in `bpmodels` with the `chain_sim()` function. 

## Disease data

Let's create an outbreak dataset, assuming the cases are sampled from a negative binomial
distribution with mean = 5 and dispersion = 7.5. These parameter values are arbitrarily
chosen for illustrative purposes.

```{r data_generation, message=FALSE}
set.seed(12)
cases_df <- data.frame(date = as.Date('2023-01-01') + seq_len(12),
                       cases = rnbinom(12, size = 7.5, mu = 5)
                       )
head(cases_df)
ggplot(cases_df, aes(x = date, y = cases)) + geom_col()
```

## Preparing the inputs  

```{r input_prep, message=FALSE}
# We will create a vector of starting times for each case, using the time of the index cases as the reference point
cases_df$days_since_index <- as.integer(cases_df$date - min(cases_df$date))

#'Disaggregate the time series 
case_times <- unlist(mapply(function(x, y) rep(x, times = ifelse(y == 0, 1, y)), 
                       cases_df$days_since_index, 
                       cases_df$cases
                       )
                       )
                       


#' Date to end simulation (14 day projection in this case)
projection_window <- 14 #2 week ahead projection
project_to_date <- max(cases_df$days_since_index) + projection_window 


#' Number of simulations and maximum chain size
sim_rep <- 1000
cases_to_project <- 1000


### Specifying the `serial` argument to `chain_sim()`
#' Assume serial interval follows log-normal distribution with mean, mu = 4.7, 
#' and standard deviation, sigma = 2.9, then the desired standard deviation, si_sd, 
#' and mean, si_mean, are
sigma = 2.9
mu = 4.7

si_sd <- sqrt(log(1 + (sigma/mu)^2)) #log standard deviation
si_mean <- log((mu^2)/(sqrt(sigma^2 + mu^2))) #log mean

#' serial interval function
serial_interval <- function(sample_size) {
  si <- rlnorm(sample_size, meanlog = si_mean, sdlog = si_sd)
  return(si)
}
```

## Simulations
```{r simulations, message=FALSE}
## Chain log-likelihood simulation
sim_chain_sizes <- lapply(seq_len(sim_rep),
                           function(sim){chain_sim(
                               n = length(case_times),
                               offspring = "nbinom",
                               mu = 2.0,
                               size = 0.38,
                               stat = "size",
                               infinite = cases_to_project,
                               serial = serial_interval,
                               t0 = case_times,
                               tf = project_to_date,
                               tree = TRUE
                           ) |> 
                               mutate(sim = sim)} 
                          )

sim_output <- do.call(rbind, sim_chain_sizes) 
```


### Post-processing
```{r post_processing}
ref_date <- min(cases_df$date)

incidence_ts <- sim_output |> 
  mutate(day = floor(time)) |> 
  group_by(sim, day) |> 
  summarise(cases = n()) |>  
  ungroup()


## Median cases by date.  
median_daily_cases <- incidence_ts |>
  group_by(day)|>
  summarise(median_cases = median(cases)) |>
  ungroup()|>
  arrange(day) |>
  mutate(date = ymd(ref_date) + 0:(project_to_date - 1))

```


## Visualization
```{r visualisation}
# Visualization
cases_plot <- ggplot(data = median_daily_cases) +
  geom_col(aes(x = date, y = median_cases),
           fill = "tomato3",
           size = 1
  ) +
  scale_y_continuous(breaks = seq(0, max(median_daily_cases$median_cases) + 20, 20),
                     labels = seq(0, max(median_daily_cases$median_cases) + 20, 20)
  ) +
  labs(x = 'Date', y = 'Daily cases (median)') + 
  theme_minimal(base_size = 14)

print(cases_plot)
```

